### 2024-04-10 docs/SOSP 2024/intro.tex
Edit intro

%Despite these recent advancements, one dominant family of models, namely decision forest models have not seen the same level of acceleration.
Decision forest models remain the mainstay for machine learning over tabular data~\cite{DLNotAllYouNeed,TreebasedOutperformDL}. 
Their robustness, interpretability, and ability to handle missing data make them a popular choice for a wide range of applications. 
\TODO{a few more lines, classification, regression, etc.}
%We observe that while scalable performance on neural network models has received significant attention, decision tree inference has not seen the same level of acceleration.
A recent survey~\cite{kaggle}\dots. 
The survey also finds that, the cost of inference is the most critical factor in the overall cost of deploying a machine learning model.
This is because, in production settings, each model is trained once and often used for inference a few thousand times. 
%

This paper is motivated by the need to accelerate decision tree inference to achieve portable performance on a variety of client hardware.
In particular, we focus on a range of commodity CPUs and low-to-mid-range GPUs, a class of hardware that has seen widespread adoption across client/edge devices used for inference.

Decision forest models are composed of a large collection of decision trees (100-1000), and inference involves traversing down each tree in the forest and aggregating the predictions. Inference is typically done in a batched setting, where multiple inputs are processed simultaneously.

%Despite the simplicity of the model, decision tree inference is computationally expensive, and is surprisingly hard to scale to a large number of threads
Despite the simplicity of the model and the availability of multiple sources of coarse grain parallelism (parallelism across inputs in a batch and parallelism across trees), existing systems do not consistently scale well across different models. 
This is because existing systems exploit a limited set of optimizations and often specialize the implementation to a specific hardware platform, and this limits their portability.

%\TODO{cover Rapids, Tahoe, XGBoost and Treebeard}.  
Evaluation on a diverse set of models highlights that the best implementation often requires a careful combination of many optimization strategies. For example, as we simultaneously traverse multiple trees, managing the working set of nodes is critical to get scalable performance. This requires a combination of techniques like data layout optimizations, loop transformations, and memory access optimizations. Existing systems each use specific data layouts for the underlying tree (XGBoost~\cite{XGBoost} uses a sparse representation, RAPIDs~\cite{Rapids} FIL~\cite{FTL} uses what is called the reorg representation 
and Tahoe~\cite{Tahoe} uses a variation of the reorg representation) and propose simple loop transformations (XGBoost uses \dots() \TODO{complete}) that together determine the accesses patterns and the working set. They do not perform consistently well across different models and hardware platforms.
Managing the working set is just one of several optimizations that are critical for high-performance decision tree inference.
 \TODO{Improve/drop the next line}
 Others include tree ordering,  incremental reduction of predicted values, interleaving across multiple trees, and so on.

>>>>

%Despite these recent advancements, one dominant family of models, namely decision forest models have not seen the same level of acceleration.
Decision forest models remain the mainstay for machine learning over tabular data~\cite{DLNotAllYouNeed,TreebasedOutperformDL}. 
Their robustness, interpretability, and ability to handle missing data make them a popular choice for a wide range of applications. 
\TODO{a few more lines, classification, regression, etc.}
%We observe that while scalable performance on neural network models has received significant attention, decision tree inference has not seen the same level of acceleration.
A recent survey~\cite{kaggle}\dots. 
The survey also finds that, the cost of inference is the most critical factor in the overall cost of deploying a machine learning model.
This is because, in production settings, each model is trained once and often used for inference millions of times. 
However, inference is run on a variety of hardware platforms, ranging from low to high-end CPUs and GPUs. 
This paper is motivated by the need to accelerate decision tree inference to achieve portable performance on a variety of hardware.
% In particular, we focus on a range of commodity CPUs and GPUs, a class of hardware that has seen widespread 
% adoption across client/edge devices used for inference.

Decision forest models are composed of a large collection of decision trees (100-1000), and inference involves 
traversing down each tree in the forest and aggregating the predictions. Inference is typically done in a batched
setting, where multiple inputs are processed simultaneously.
Despite the simplicity of the model and the availability of multiple sources of coarse grain parallelism (parallelism 
across inputs in a batch and parallelism across trees), existing systems do not consistently scale well across different
models even on the limited set of targets they support. 
This is because existing systems exploit a limited set of optimizations and often specialize the implementation 
to a specific hardware platform, and this limits their portability.

%\TODO{cover Rapids, Tahoe, XGBoost and Treebeard}.  
Evaluation on a diverse set of models highlights that the best implementation often requires a careful combination 
of many optimization strategies. For example, simultaneously traversing multiple trees with many threads, while managing the working set is 
critical to get scalable performance. This requires a combination of techniques like data layout optimizations, 
loop transformations, and memory access optimizations. Existing systems each use specific data layouts for the underlying
tree and use fixed loop structures. 
Parallelism, loop structure and model representation together determine the accesses patterns and the working set. A fixed configuration 
as used by existing systems does not perform consistently well across different models and hardware platforms.
XGBoost~\cite{XGBoost} uses a sparse representation for the model and a loop structure that processes one tree for a block of rows before moving to 
the next tree. RAPIDS FIL~\cite{FIL} uses a reorg representation and partitions trees across a fixed number of threads. Tahoe~\cite{Tahoe} uses a
variation of the reorg representation and has four predefined inference strategies from which it picks one based on an analytical model.
These systems are all library based and are hard to extend. 
\TreebeardOLD{}, the state-of-the-art model compiler for CPUs, supports two fixed loop structures and does not scale well with increasing number of threads.
